HO unification from object language to meta language

### 2024-02-15 main.tex
update tex

\def\ol{\acronym{ol}} % object language
\def\ml{\acronym{ml}} % meta language
\def\lf{\acronym{lf}} % logical framework
\def\ho{\acronym{ho}} % higher order

@@

% syntax of coq terms in elpi
In particular, \coq terms are encoded following the following (very simplified)
data type

@@

\begin{tcolorboxSimple}
  \begin{tabular}{c|c}
    \begin{minipage}[c]{0.50\linewidth}
      \begin{elpicode}
        kind term type.
        type const gref -> term.
        type app list term -> term.
        type fun (term -> term) -> term.
        type prod (term -> term) -> term.
      \end{elpicode}
    \end{minipage}
    &
    \begin{minipage}[c]{0.48\linewidth}
      \begin{textcode}
        Example:
        a       ==> const 0
        f a     ==> app[const `f, const `a]
        ~$\lambda$~x => x ==> fun x\ x
        ~$\forall$x, F x~ ==> prod x\ app[F, x]
      \end{textcode}
    \end{minipage}
  \end{tabular}
\end{tcolorboxSimple}

\noindent where a \coq constant is a gref\footnote{a gref is the opaque type for \coq
identifier in \elpi} inside the \elpiIn{const} constructor, the \coq application
is embedded in the \elpiIn{app} node, having the head and the arguments of the
application expressed as a list of terms. Lambda abstractions are translated
into the node \elpiIn{fun} binding a term to another term. Finally forall
quantification on the for \coqIn{~$\forall$~x,F} into \elpiIn{forall x\ F}.

@@

Let's take the following example\footnote{taken from the \stdpp library}:
\begin{tcolorboxSimple}
  \begin{tabular}{c|c}
    \begin{minipage}[c]{0.50\linewidth}
      \begin{coqcode}
        Instance forall_dec A (P: A ~$\to$~ Prop):
          Finite A ~$\to$~ 
          ~$\forall$~x, Decision (P x) ~$\to$~
          Decision (~$\forall$~x, P x).
      \end{coqcode}
    \end{minipage}
    &
    \begin{minipage}[c]{0.46\linewidth}
      \begin{elpicode}
        tc-Decision (prod x\ app[P, x]) :- 
          % premise for Finite A
          % premise for $\forall$x, Decision (P x)
      \end{elpicode}
    \end{minipage}
  \end{tabular}
\end{tcolorboxSimple}

The \coq instance on the left is compiled in the \elpi rule on the right. If the 
query to solve (already in the \elpi syntax) is

\elpiInEq{tc-Decision (prod x\ const `nat)}

\noindent it is clear that the \elpi rule above will not unify with the query,
Since the rigid node \elpiIn{app} is different from the rigid node
\elpiIn{const}. The problem come form the lost of the fact that \coqIn{P} is an
higher-order variable in \coq, and in \elpi it is translated into a first order
variable (the variable \elpiIn{P} in \elpi does not see the binder \elpiIn{x}).

Another unification problem we can da notare che il problema dopo, is the FO
approx. Coq applica la regola forall\_dec sul goal (prod x\ app[f, y, x]), ma forse 
bisognerebbe vedere se altri solver accettano questa euristica. Così facciamo 
vedere che non siamo i soli a farlo. Nel caso contrario bisogna dire come funziona
l'algo di unif di coq in questo caso specifico

\elpiInEq{tc-Decision (prod x\ app[const `f, const `y, x])}

Specificare anche che non riusciamo a fare ne eta ne beta riduzione.

Dire che elpi unifica in HO e con eta-beta

>>>>

\def\ol{\acronym{ol}} % object language
\def\ml{\acronym{ml}} % meta language
\def\lf{\acronym{lf}} % logical framework
\def\ho{\acronym{ho}} % higher order
\def\Forall{$\forall$}

@@

% syntax of coq terms in elpi
In particular, \coq terms are encoded with the following (very simplified)
data type

@@

\noindent %
\begin{tabular}{c|c}
  \begin{minipage}[c]{0.50\linewidth}
    \begin{elpicode}
      kind term type.
      type const gref -> term.
      type app list term -> term.
      type fun (term -> term) -> term.
      type prod (term -> term) -> term.
    \end{elpicode}
  \end{minipage}
  &
  \begin{minipage}[c]{0.48\linewidth}
    \begin{textcode}
      Example:
      a       ==> const `a
      f a     ==> app[const `f, const `a]
      ~$\lambda$~x => x ==> fun x\ x
      ~$\forall$x, F x~ ==> prod x\ app[F, x]
    \end{textcode}
  \end{minipage}
\end{tabular}

\noindent where a \coq constant is a gref\footnote{a gref is the opaque type for \coq
identifier in \elpi} inside the \elpiIn{const} constructor, the \coq application
is embedded in the \elpiIn{app} node, having the head and the arguments of the
application expressed as a list of terms. Lambda abstractions are translated
into the node \elpiIn{fun} binding a term to another term. Finally forall
quantification on the for \coqIn{~$\forall$~x,F} into \elpiIn{prod x\ F}.

@@

Let's take the following example from the \stdpp library:

\noindent %
\begin{tabular}{c|c}
  \begin{minipage}[c]{0.50\linewidth}
    \begin{coqcode}
      Instance forall_dec A (P: A ~$\to$~ Prop):
        Finite A ~$\to$~ 
        ~$\forall$~x, Decision (P x) ~$\to$~
        Decision (~$\forall$~x, P x).
    \end{coqcode}
  \end{minipage}
  &
  \begin{minipage}[c]{0.46\linewidth}
    \begin{elpicode}
      tc-Decision (prod x\ app[P, x]) :- 
        % premise for Finite A
        % premise for $\forall$x, Decision (P x)
    \end{elpicode}
  \end{minipage}
\end{tabular}

\noindent In our type-class solver, we compile the \coqIn{forall_dec} instance
of \coq (on the left) into the \elpi rule on the right. This compilation, which
is blindly built from the type of \coqIn{forall_dec}, hides some unification
properties accepted by the unification algorithm of the \ol but that are
rejected by the unification algorithm of the \ml. The main problem of this
compilation is how the subterm \coqIn{~$\forall$~x, P x} is translated in
the \ml rule; \coqIn{P} is a higher order variable, in the former subterm, it
does see the binder \coqIn{x}. However, the rule in the \ml, consider a first
order variable \elpiIn{P}, not seeing \elpiIn{x}. Furthermore, this subterm is
lodged within the \elpiIn{app} constructor making unification very structural.

% Esempio HO in PF
\elpiInEq{tc-Decision (prod x\ const `nat)}%
% Esempio FO align
\elpiInEq{tc-Decision (prod x\ app[const `f, const `y, x])}

\noindent The two example of queries above, already in the syntax of the \ml,
represent two terms that would unify in the \ol, but fail in the \ml. In
particular, the first fails since the \ml is not able to unify \elpiIn{const
`nat} with the term \elpiIn{app[P, x]}: they have different rigid heads.

The second goal refers to an approximation of the unification\todo{maybe there's
something to cite} algorithm of the \ol. Spiegare l'approssimazione. E che
attraverso il nostro framework è possibile "spiegare" l'algorithm di unification
usato da ogni \ol nel ml desiderato. \todo{something to do here...}

Following the definition give just before, the terms \elpiIn{app[const `f, const
`y, x]} and \elpiIn{app[P, x]} should unify with substitution $\theta := \{P
\mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
is not possible in the \ml since, this time we have two terms with the same head
(the \elpiIn{app} node), but the corresponding lists have different lengths.

An immediate solution to solve these unification problems\todo{mettere da qualche parte anche eta e beta}
would be to be lazy and build very general rule for each instance in the database.
This would build a rule with the following shape:

\begin{center}
  \begin{minipage}[c]{0.46\linewidth}
    \begin{elpicode}
      tc-Decision X :- 
        ol.unify X (prod x\ app[P, x]),
        % premise for Finite A
        % premise for $\forall$x, Decision (P x)
    \end{elpicode}
  \end{minipage}
\end{center}

\noindent In this second implementation, any call to the \elpiIn{tc-Decision}
will unify with the head of the rule for \coqIn{forall_dec}. The the first
premise will ensure that the arguments unify with the what we expect from the
instance definition. Notice that, this time, the very unification algorithm of
the \ol is exploited, overcoming the unification problems listed just before.
This approach is quite appealing but it has two main disadvantages: the first
lays on the fact that the communication between the \ol and the \ml can
potentially be slow, this is due to the transformation of terms from one language
to the other. \todo{here i refer to constr->elpi e elpi->constr}
The other drawback of this latter encoding is that clause discrimination cannot
be done on the head of the clause, for instance, any indexing 
algorithm allowing to filter rules become suddenly useless.

% Contribution
Finally

% da notare che il problema dopo, is the FO
% approx. Coq applica la regola forall\_dec sul goal (prod x\ app[f, y, x]), ma forse 
% bisognerebbe vedere se altri solver accettano questa euristica. Così facciamo 
% vedere che non siamo i soli a farlo. Nel caso contrario bisogna dire come funziona
% l'algo di unif di coq in questo caso specifico
